Mapping Cancer: the Potential of Cartograms and Alternative Map Displays

### 2020-05-26 literature.Rmd
lots of fixes in both word and rmd file and changes to code for fig 2

Cancer atlases communicate cancer statistics over geographic domains, typically with a choropleth map. A choropleth map subdivides these domains into administrative regions such as countries, states, or suburbs. When communicating human-related statistics, the choropleth has a disadvantage in that it draws attention to sparsely populated rural areas to the neglect of small inner - city areas. The smaller geographic areas are important to consider if they are densely populated. Alternative map displays, such as a cartogram or a hexagon tile map, can shift the attention of map users from the large rural areas by decreasing their size on the map display. This means alternative displays can be more effective at accurately communicating spatial patterns across spatial areas. This study summarises current practices for cancer atlases and investigates the alternative map displays that could be used to accurately represent the distribution of cancer statistics across a population, as many cancer atlases lack appropriate displays for population statistics. It is recommended that alternative displays are included in cancer atlases for a perceptually accurate display of the distribution of the burden of cancer over the population. If it possible given the time and budget constraints, atlases should also present countries or regions containing hetergeneous spatial areas in an alternative display, in addition to the familiar choropleth map. With the ease of today’s technology, user interaction with the displays is also encouraged. Users should also be able to interactively display different statistics, such as incidence rate or relative incidence, or filter by demographic variables.

# 1. Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Researchers, health authorities, governments, not-for-profits and the media are common communicators of cancer statistics. They often present statistics to the public as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations for the geographical units, especially for privacy protection, but also for political and policy purposes. Examples of typical geographical units include states, provinces, local government areas, and post/zip codes. Counts and incidence rates are often obtained by state health registries. For this data privacy and ethics are key concerns, even when combined to small area statistics.
This type of data is routinely collected for public health reasons and may be made available to the general public as a service to the community.

<!-- # Cancer atlases -->
<!-- Displaying of cancer data on choropleth maps -->
A cancer atlas is a map, or collection of maps, commoly representing cancer incidence or mortality patterns across a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, and geographic correlations [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization. Cancer maps and atlases date back to Haviland's maps in 1875, with more modern atlases directly evolving from early work in US cancer atlases, appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power and the availability of geographic information systems software [@SE].

@@

The paper is structured as follows. The next section introduces cancer atlases and the worldwide implementations. Choropleths are the common approach to disease maps and presents examples of atlases in use today and discusses the limitations of the choropleth map. Section 3 describes alternative displays, including the cartogram, which is useful when the map has heterogeneously sized geographic units. Section 4 presents the limitations of the production and use of alternative displays. Disease maps are more useful when made interactive, and common options are described in Section 5, along with a discussion of benefits and disadvantages.

@@

The measures displayed are typically aggregations of data values over small areas or model estimates. This to protect patient privacy, and for numerical stability. The counts of cases in small areas can be difficult to obtain as they are often protected for privacy reasons. The information released and the statistics presented in worldwide cancer atlases are often model estimates. The measures described in Table 1 use incidence as the statistic of interest, which reports the number of new cases. These same statistics can be used to describe the mortality, that is, Standardised Incidence Rate is analogous to the Standarised Mortality Rate.

@@

| Fig. | Atlas | Data | Statistic | Source |
|----|-----------|-------------|-----------|--------------|
| a | The Environment and Health<br> Atlas of England and Wales<br> [@EnvEnglandWales2] | Sex: Women<br> Year: 2010<br> Cancer: Lung | Relative risk | Office for National Statistics<br> (ONS) (England) and from the<br> Welsh Cancer Intelligence and<br> Surveillance Unit (WCISU) [1a](http://www.envhealthatlas.co.uk/homepage/) |
| b | Globocan 2018: Estimated Cancer<br> Incidence, Mortality and <br> <br> prevalence Worldwide<br> [@Globocan] | Sex: Both<br> Year: 2018<br> Cancer: All invasive | Age standardized<br> incidence rates<br> (per 100,000) | World Health Organization's<br> International Agency for<br> Research on Cancer [1b](https://www.iarc.fr/<br>)  |
| c | Atlas of Cancer in Queensland<br>  [@QLDcancerAtlas] | Sex: Males<br> Year: 1998 - 2007<br> Cancer: Lung | Relative incidence<br> ratio | Queensland Cancer Registry [1c](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/) |
| d | Australian Cancer Atlas<br> [@ACA] | Sex: Males<br> Year: 2010 - 2014<br> Cancer: Lung | Average standardised<br> incidence rate | Australian Cancer Atlas [1d](https://atlas.cancer.org.au/) |
| e | United States Cancer<br> Statistics: An Interactive<br> Cancer Statistics Website<br>  | Sex: Both<br> Year: 2016<br> Cancer: All types | Incidence rate per<br> 100,000 | With data from state cancer<br> registries: <br> Centers for Disease Control and Prevention [1e](https://www.cdc.gov/) |
| f | Map of Cancer Mortality<br> Rates in Spain<br> [@cancerSpain] | Sex: Both<br> Year: 2004 - 2008<br> Cancer: All types | Relative risk | <br> Map of cancer mortality rates<br> in Spain [1f](https://elpais.com/elpais/2014/10/06/media/1412612722_141933.html) |
| g | Atlas of Childhood<br> Cancer in Ontario<br> [@OntarioPediatric] | Sex: Both<br> Year: 1995 - 2004<br> Cancer: All types | Incidence rate per<br> 100,00 | <br> The Pediatric Oncology Group<br> of Ontario Networked Information<br> System [1g](https://www.pogo.ca/research-data/pogonis-childhood-cancer-database/data-requests/) |

@@

There are two key issues to consider when creating alternative map displays, (1) the intended audience of the map, and (2) its purpose. Nusrat and Kobourov [@SAIC] provided a framework to investigate implementations of the many algorithms presented, and the "statistical accuracy, geographical accuracy, and topological accuracy". 
The alternative map displays in Figure 2 a - c are created by resizing and reshaping the states of the USA to match the 2015 population of the state. This provides a better sense of the extent of disease relative to the population in the country and can help prevent overlooking the impact of the disease within physically small but population-dense states. Map creators give each state equal size and thus equal emphasis in Figure 2d the hexagon tile map.

@@

## 4.1 Performance of choropleth maps or cartograms for Australia

@@

Monmonier [@HTLWM] recommends using interactivityto allow users to explore the map for more information and provide flexibility for the display. The user can toggle between different variables, map views or even multiple realizations of future scenarios [@goodchild1994introduction]. This provides additional mechanisms for the users to digest the uncertainty of the available information [@maceachren1992visualizing, @van1994visualization]. When the needs of the audience are changeable and are also the priority, the map creator can allow interactivity for map users to explore a data set through dynamic interactions. This can allow inspection of the data from many views [@DQBCM]. User interaction with maps helps to understand and interpret the spatial distribution of disease, to validate, explain or explore the presented statistics and their relationships to each other [@TNTEA].

@@

<!-- summary of mapping -->
This paper provides an overview of mapping practices as commonly used for cancer atlases. The conventional approach is the choropleth map, and it is widely used. The choropleth map suffers when there are small geographic units, as occurs in Australia where the population is concentrated on the coast, the information about the burden of cancer on those communities can be hidden. Making an inset can clarify congested regions but this breaks the viewers' attention as they shift focus from the map to the inset, and if there are many congested areas, many insets would be needed. The map alternatives implement trade-offs between the familiar shapes, and the importance of the geographic areas in the context of the areas. Given the population or a cancer statistic for each area, the geographic size or shape will change. Alternative displays allow the spatial distribution of cancer data to be digested by map users.

@@

<!-- statistics -->
Many statistics are commonly used in cancer displays. It is common to see incidence rates, or relative rates which measure how far a region is above or below the average. The purpose of using an age standardised relative rate is, perhaps the desire to pinpoint the areas that need attention because they have higher than expected rates. 
<!--However, we lose the incidence rate information and thus interpretability. --> A region might be much higher than average, but it may not be close to a health concern, because all regions have a low incidence. Supplementary materials can allow map users to recognise when this occurs.

<!-- interaction -->
Interaction with maps is an important component of public atlases, and is becoming increasingly straightforward to add with today's technology. The purpose of interaction in public atlasses is to provide access to more information than is possible to display in a single map, without overwhelming the viewer. Too many choices can similarly overwhelm a viewer, and thus decisions do need to be made about content to provide for accurate and comprehensive communication of information. Similarly, providing ways for users to interact with the display encourages engagement, and creative, efficient, elegant, interactive tools elicit curiosity about the data.

>>>>

Cancer atlases communicate cancer statistics over geographic domains, typically with a choropleth map. A choropleth map subdivides these domains into administrative regions such as countries, states, or suburbs. When communicating human-related statistics, the choropleth has a disadvantage in that it draws attention to sparsely populated rural areas to the neglect of small inner - city areas. The smaller geographic areas are important to consider if they are densely populated. Alternative map displays, such as a cartogram or a hexagon tile map, can shift the attention of map users from the large rural areas by decreasing their size on the map display. This means alternative displays can be more effective at accurately communicating spatial patterns across spatial areas. This study summarises current practices for cancer atlases and investigates the alternative map displays that could be used to accurately represent the distribution of cancer statistics across a population, as many cancer atlases lack appropriate displays for population statistics. It is recommended that alternative displays are included in cancer atlases for a perceptually accurate display of the distribution of the burden of cancer over the population, in addition to the familiar choropleth map, if possible given time and budget constraints.  With the ease of today’s technology, user interaction with the displays is also encouraged. Users should also be able to interactively display different statistics, such as incidence rate or relative incidence, or filter by demographic variables.

# 1. Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Researchers, health authorities, governments, not-for-profits and the media are common communicators of cancer statistics. They often present statistics to the public as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations for the geographical units, especially for privacy protection, but also for political and policy purposes. Examples of typical geographical units include states, provinces, local government areas, and post/zip codes. Counts and incidence rates are often obtained by state health registries. For this data, privacy and ethics are key concerns, even when combined to small area statistics.
This type of data is routinely collected for public health reasons and may be made available to the general public as a service to the community.

<!-- # Cancer atlases -->
<!-- Displaying of cancer data on choropleth maps -->
A cancer atlas is a map, or collection of maps, commonly representing cancer incidence or mortality patterns across a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, and geographic correlations [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization. Cancer maps and atlases date back to Haviland's maps in 1875, with more modern atlases directly evolving from early work in US cancer atlases, appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power and the availability of geographic information systems software [@SE].

@@

The paper is structured as follows. The next section describes the common statistics displayed in disease mapping. Section 3 focuses on disease map visualizations. It describes cancer atlases, presents examples of atlases in use today. It discusses the limitations of the most commonly used technique for disease mapping, the choropleth map. This section also describes alternative displays, including the cartogram, which are useful when the map has heterogeneously sized geographic units. Section 4 presents the limitations in the production and use of alternative displays. Disease maps are more useful when made interactive, and common options are described in Section 5, along with a discussion of benefits and disadvantages. Section 6 summarizes this survey of the literature, and provides recommendations.

@@

The measures displayed are typically aggregations of data values over small areas or model estimates. This to protect patient privacy, and for numerical stability. The counts of cases in small areas can be difficult to obtain as they are often protected for privacy reasons. The information released and the statistics presented in worldwide cancer atlases are often model estimates. The measures described in Table 1 use incidence as the statistic of interest, which reports the number of new cases. These same statistics can be used to describe the mortality, that is, standardized Incidence Rate is analogous to the Standardized Mortality Rate.

@@

| Fig. | Atlas | Data | Statistic | Source |
|----|-----------|-------------|-----------|--------------|
| a | The Environment and Health<br> Atlas of England and Wales<br> [@EnvEnglandWales2] | Sex: Women<br> Year: 2010<br> Cancer: Lung | Relative risk | Office for National Statistics<br> (ONS) (England) and from the<br> Welsh Cancer Intelligence and<br> Surveillance Unit (WCISU) [1a](http://www.envhealthatlas.co.uk/homepage/) |
| b | Globocan 2018: Estimated Cancer<br> Incidence, Mortality and <br> <br> prevalence Worldwide<br> [@Globocan] | Sex: Both<br> Year: 2018<br> Cancer: All invasive | Age standardized<br> incidence rates<br> (per 100,000) | World Health Organization's<br> International Agency for<br> Research on Cancer [1b](https://www.iarc.fr/<br>)  |
| c | Atlas of Cancer in Queensland<br>  [@QLDcancerAtlas] | Sex: Males<br> Year: 1998 - 2007<br> Cancer: Lung | Relative incidence<br> ratio | Queensland Cancer Registry [1c](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/) |
| d | Australian Cancer Atlas<br> [@ACA] | Sex: Males<br> Year: 2010 - 2014<br> Cancer: Lung | Average standardized<br> incidence rate | Australian Cancer Atlas [1d](https://atlas.cancer.org.au/) |
| e | United States Cancer<br> Statistics: An Interactive<br> Cancer Statistics Website<br>  | Sex: Both<br> Year: 2016<br> Cancer: All types | Incidence rate per<br> 100,000 | With data from state cancer<br> registries: <br> Centers for Disease Control and Prevention [1e](https://www.cdc.gov/) |
| f | Map of Cancer Mortality<br> Rates in Spain<br> [@cancerSpain] | Sex: Both<br> Year: 2004 - 2008<br> Cancer: All types | Relative risk | <br> Map of cancer mortality rates<br> in Spain [1f](https://elpais.com/elpais/2014/10/06/media/1412612722_141933.html) |
| g | Atlas of Childhood<br> Cancer in Ontario<br> [@OntarioPediatric] | Sex: Both<br> Year: 1995 - 2004<br> Cancer: All types | Incidence rate per<br> 100,00 | <br> The Pediatric Oncology Group<br> of Ontario Networked Information<br> System [1g](https://www.pogo.ca/research-data/pogonis-childhood-cancer-database/data-requests/) |

@@

The intended audience and the communication purpose are important to consider when creating alternative map displays.  Nusrat and Kobourov [@SAIC] provided a framework to investigate implementations of the many algorithms presented, and the "statistical accuracy, geographical accuracy, and topological accuracy". 
The alternative map displays in Figure 2 a - c are created by resizing and reshaping the states of the USA to match the 2015 population of the state. This provides a better sense of the extent of disease relative to the population in the country and can help prevent overlooking the impact of the disease within physically small but population-dense states. Map creators give each state equal size and thus equal emphasis in Figure 2d the hexagon tile map.

@@

## 4.1 Performance of cartograms for Australia

@@

Monmonier [@HTLWM] recommends using interactivity to allow users to explore the map for more information and provide flexibility for the display. The user can toggle between different variables, map views or even multiple realizations of future scenarios [@goodchild1994introduction]. This provides additional mechanisms for the users to digest the uncertainty of the available information [@maceachren1992visualizing, @van1994visualization]. When the needs of the audience are changeable and are also the priority, the map creator can allow interactivity for map users to explore a data set through dynamic interactions. This can allow inspection of the data from many views [@DQBCM]. User interaction with maps helps to understand and interpret the spatial distribution of disease, to validate, explain or explore the presented statistics and their relationships to each other [@TNTEA].

@@

<!-- summary of mapping -->
This paper provides an overview of mapping practices as commonly used for cancer atlases. The conventional approach is the choropleth map, and it is widely used. The choropleth map suffers when there are small geographic units, as occurs in Australia.  The population of Australia is concentrated in small areas on the coast, and a choropleth map can hide information about the burden of cancer on those communities. Making an inset can clarify congested regions but this breaks the viewers’ attention as they shift focus from the map to the inset, and if there are many congested areas, many insets would be needed. The map alternatives implement trade-offs between the familiar shapes, and the importance of the geographic areas. Given the population or a cancer statistic for each area, the geographic size or shape will change. Alternative displays allow the spatial distribution of cancer data to be digested by map users.

@@

<!-- statistics -->
Many statistics are commonly used in cancer displays. It is common to see incidence rates, or relative rates which measure how far a region is above or below the average. The purpose of using an age standardized relative rate is, perhaps the desire to pinpoint the areas that need attention because they have higher than expected rates. 
<!--However, we lose the incidence rate information and thus interpretability. --> A region might be much higher than average, but it may not be close to a health concern, because all regions have a low incidence. Supplementary materials can allow map users to recognise when this occurs.

<!-- interaction -->
Interaction with maps is an important component of public atlases, and is becoming increasingly straightforward to add with today's technology. The purpose of interaction in public atlases is to provide access to more information than is possible to display in a single map, without overwhelming the viewer. Too many choices can similarly overwhelm a viewer, and thus decisions do need to be made about content to provide for accurate and comprehensive communication of information. Similarly, providing ways for users to interact with the display encourages engagement, and creative, efficient, elegant, interactive tools elicit curiosity about the data.
